OpenMPIR

### 2017-09-08 paper.tex
Won't worry about openmp task popularity for now
rather shallow editing

\title{OpenMPIR}
\subtitle{Implementing OpenMP tasks with Tapir}
\author{George Stelle}
\affiliation{\institution{Los Alamos National Laboratory}}
\email{stelleg@lanl.gov}

@@

\begin{abstract}
Optimizing compilers for task-level parallelism are still in their infancy.
This work explores a frontend that compiles OpenMP tasking semantics to Tapir,
an extension to LLVM IR that represents fork-join parallelism. This enables
analyses and optimizations that were previously inaccessible to OpenMP codes,
as well as the ability to target additional runtimes at code generation. Using
a Cilk runtime backend, we compare results to existing OpenMP implementations.
Initial performance results for the Barcelona OpenMP tasking suite and show
performance improvements over these existing implementations.
\end{abstract}

@@

When writing task-parallel programs today, one has a large selection of
potential programming models and implementations to consider. Unfortunately, 
despite having some significant overlaps in semantics, parallel programming
models such as OpenMP~\cite{openmp}, Cilk~\cite{cilk}, Kokkos~\cite{kokkos},
HPX~\cite{hpx}, Charm++~\cite{charm}, Qthreads~\cite{qthreads},
pthreads\cite{pthreads}, MPI~\cite{mpi}, Chapel~\cite{chapel}, UPC~\cite{upc},
OpenCL~\cite{opencl}, OpenACC~\cite{openacc} among others have limited ability to
interoperate either at the level of compile-time analysis or at run-time
execution. In addition, many of these programming models are implemented as
runtime libraries with only primitive compiler support. As a result, most are
not able to take advantage of the compiler's analysis and optimization
capabilities.  While there has been some recent work on specializing
compilers to reason about parallel programming libraries~\cite{Moss_2016},
fragmentation among models presents a significant challenge for compiler
writers, which must choose a single model, or multiple intermediate forms, to
use for analysis and optimization. An ideal solution would be a form of
common representation of parallel semantics and constructs.

@@

Compilers implement the majority of their analyses and optimizations on
intermediate representations \textbf{(IRs)} which allow such transformations
to be written once with relative ease and apply to a variety of source
front-end languages such as C or C++.  Historically, the IR of mainstream
compilers such as LLVM~\cite{lattner2004llvm} or GCC's
Gimple~\cite{merrill2003generic} haven't suppored parallel constructs. As a
consequence, compilers utilizing these IRs haven't had the ability to reason
about parallelism. The recent work of Schardl et al.~\cite{tapir} has shown
that the LLVM IR can be extended to represent fork-join parallelism without
requiring a major rewrite.  They did so by extending the LLVM instruction set
with three instructions capable of representing fork-join parallelism. To
benchmark performance and test accuracy, they wrote a frontend which translated
programs written with Cilk, a parallel programming model for C/C++, into the
Tapir IR. They also implemented a backend to \textbf{lower} Tapir programs to
vanilla LLVM IR with embedded Cilk runtime calls. They suggested that a similar
approach could be taken with OpenMP and other such frameworks.

In this work, we put that suggestion to the test, implementing OpenMP tasks by
compiling them to Tapir instructions. This allows us to both analyze and optimize
OpenMP programs as well as compile programs written in OpenMP to use other runtimes.
For this paper, our prototype implementation uses the Cilk runtime as a backend,
and we run our prototype implementation on the Barcelona OpenMP task suite~\cite{barcelona}. 
We discuss the kinds of optimizations this enables, and how
the work can be extended to include other parallelism constructs and semantics
in OpenMP. We also discuss the possibility of extending this work to other
programming models, which would help to alleviate some of the fragmentation
issues discussed above.

The contributions of this work include:

\begin{itemize}
  \item A prototype frontend that transforms OpenMP task constructs to Tapir IR.
  \item Benchmarks of the implementation using a Cilk runtime backend, showing
        performance improvements over existing OpenMP implementations.
  \item An initial investigation into the reasons for these performance improvements.
\end{itemize}

Our hope for this paper is to move the community towards further conversations on a 
shared IR that can be used for many programming models. The work presented in this paper only
represents a small step in that direction and there is still a herculean amount
of design and implementaton details to be considered. We argue that the benefits vastly outweigh 
the costs and we hope to see additional future work explore additional possibilities.

@@

The remainder of the paper is organized as follows: In Section~\ref{Sec:Background} 
we give background and motivation for the paper. Following that, we describe OpenMP
tasking in Section~\ref{Sec:OpenMP}, and Tapir in Section~\ref{Sec:Tapir}. We then 
discuss the implementation, and how we compile OpenMP task constructs to Tapir IR in
Section~\ref{Sec:Implementation}. In Section~\ref{Sec:Evaluation} we describe the 
evaluation setup, including what implementations we compare to and how. We discuss
the results of evaluation in Section~\ref{Sec:Results}, including possible
explanations for discrepancies. We discuss the implications of the results ans the
significance of this work in a large context in~\ref{Sec:Discussion}. Finally,
we cover some of the many ways the work can be extended in
Section~\ref{Sec:Future}, and conclude in Section~\ref{Sec:Conclusion}.

@@

LLVM has gained traction in the broad community for its simplicity,
relative ease of both experimentation and modifications, and extensibility~\cite{lattner2004llvm}.
Historically, LLVM has only had a sequential semantics: there is no explicit notion
of concurrent or parallel processes. This has resulted in existing compilers
treating parallelism constructs as thin wrappers for calls into runtime
systems. These calls are infeasible for the compiler to reason about, due to
the complexity and flexibility of the runtimes being called, and the challenge
of reasoning about the corresponding semantics in a uniform fashion. The
addition of parallel constructs were recently explored by Schardl et
al.~\cite{tapir}. By adding three first-class instructions to LLVM, Tapir
enables analyses and optimizations of parallel constructs. We briefly describe
the key aspects of Tapir in Section~\ref{Sec:Tapir}.
% I removed this as the reference was already given...   PM For a more complete
% description see \cite{tapir}.

Arguably the most ubiquitous parallel programming model for on-node parallelism
in HPC, OpenMP has grown from being a way to implement parallel loops into a
large and complex specification for parallelism. This paper focuses on a 
relatively recent addition to the OpenMP specification: the \texttt{task}
and \texttt{taskwait} constructs. A detailed description of these constructs
is presented in Section~\ref{Sec:OpenMP} and a more complete description can 
be found in the OpenMP specification~\cite{openmp}.

@@

In this section we further discuss a significant challenge in HPC:
fragmentation of parallel programming models. There are many ways to write
parallel programs~\cite{openmp, qthreads, chapel, cilk, kokkos, legion, upc,
mpi} . From libraries, to language extensions, to standalone languages, to
combinations of the above, fragmentation of parallel programming models is a
problem that is only getting worse. This means that when writing tooling,
optimizations, or analyses, one has to choose which model to target as well as
understand points of contention and interoperability between them. A common
difficulty among all of these is that reasoning about parallel programs is hard
and compilers can potentially help reduce this burden. This is exactly
analogous the problem LLVM solves: compiler optimizations are hard, so sharing
them is valuable. Tapir is a demonstration that the the LLVM approach to
compilation of serial code can be effectively extended to help reason about
parallel code.

With fragmentation comes questions of compatibility, or composability, of 
different programming models. To illustrate this issue, consider the example of
a program using a library that uses OpenMP to implement parallelism internally. 
If said program uses a different programming model for parallelism, e.g. Cilk,
there is currently no good solution for ensuring that the two different backends
will cooperate with management of hardware resources.

Addressing these concerns this is a major motivation for our use of Tapir.  By
standardising an IR, in addition to easing the development of optimizations,
one avoids duplicating work across different implementations. It also enables
the possibility of multiple programming models coexisting peacefully. We'll
return to this issue later in Section~\ref{Sec:Discussion} when discussing
future work.

\section{Tapir} \label{Sec:Tapir}
Tapir is an extension to LLVM proposed by Schardl, Moses, and Leiserson
aiming to resolve issues regarding optimizing parallel code. Prior to
the introduction of Tapir, compilers would represent parallel programs
by directly translating their syntax to opaque runtime calls. This allowed
for compilers to support parallel programs, but meant that traditional 
optimizations such as code motion \cite{} or common-subexpression elimination \cite{}
weren't able to reason about parallel programs. This often led to parallel
programs running significantly slower than expected. In the Tapir project,
the authors show that it is possible to represent and optimize fork-join
parallel programs with relatively minimal modifications to the compiler: namely
the introduction of three instructions designed to interface well with an
existing compiler.

@@

Consider the first code segment defined in~\figref{search}.
Here, we find a simple OpenMP program which performs a divide-and-conquer search.
However, as written, there is a simple optimization that can be performed --
namely, that the value \code{(low+high)/2} may be computed once before either
parallel task, avoiding redundant computations. For
the serial version of this program, optimizations like this are performed
automatically. However, without the use of Tapir, these sorts of optimizations
are not run on parallel programs.

@@

\subsection{Tapir Internals}
Tapir is a compelling solution to this problem because it allows existing
analysis and optimizations to work on parallel programs without significant modifications.
To properly describe Tapir, one must briefly introduce key aspects of LLVM.
LLVM represents expression/evaluation order through a control-flow graph (CFG) of blocks that contain 
sequences of instructions. Blocks always end with terminator instructions that
define the edges of the control flow graph. For example, \figref{CFG} shows a simple
function and its corresponding LLVM code.

@@

To represent parallelism, Tapir introduces the \code{detach}, \code{reattach}, and
\code{sync} instructions. The \code{detach} statement is syntactically similar to
the conditional branch as it is also a terminator instruction with with two blocks.
However, unlike the conditional branch which denotes a choice between successors, the \code{detach} instruction denotes that its two successors may run in parallel. Semantically,
the first successor of a \code{detach} instruction is referred to as the \textbf{detached block} and represents a task that can, but is not required to, run in parallel with the
\textbf{continuation block}, or the second successor to the detach statement. The \code{detach}
is analagous to a \code{fork} in most fork-join models.

At the end of any task created by a \code{detach} instruction, one must end with a \code{reattach} instruction to the corresponding continuation block. The \code{reattach} instruction
is used to denote the end of a parallel task.

Finally, one may use the \code{sync} instruction to wait for any outstanding tasks created
by \code{detach} instructions within the current function to complete. The \code{sync}
instruction is analagous to a \code{join} in most fork-join frameworks.

@@

The initial design of the OpenMP task model~\cite{ayguade09design} established 
a basic framework for asynchronous task parallel execution in OpenMP programs.
Subsequent versions of the OpenMP specification up to the current version 4.5 
have added additional new features to the tasking model.  The \texttt{depend} 
clause codifies data dependences among tasks, indicating that a data location 
is an input or output of a task.  The runtime system ensures that a task is 
not scheduled until its input dependences are fulfilled.  The \texttt{taskloop} 
construct combines groups of independent loop iterations into explicit tasks, 
enabling composition of concurrent loop execution and independent explicit 
tasks within the same OpenMP parallel region. The \texttt{taskgroup} construct 
waits on not only all child tasks, but all descendent tasks, providing a deep 
synchronization. The \texttt{taskwait} construct allows the application to 
indicate a point at which the implementation may suspend the current task to 
work on other tasks, as may be desired for long-running tasks that generate 
many others.  The task concept was also leveraged to provide for asynchronous 
offload of data and computation to accelerators by applying the \texttt{nowait} 
clause to device constructs like the \texttt{target} construct to generate an 
asynchronous \textit{target task}.

@@

Tapir is implemented as an extension to the LLVM instruction set. Clang is a C
family compiler that has support for OpenMP extensions \cite{clang} and targets
LLVM. The existing Clang OpenMP implementation maps OpenMP constructs directly
to OpenMP runtime library calls, by wrapping a C statement in a
\texttt{CapturedStmt}. This replaces a C statement with a function call into
the OpenMP runtime library, along with a machine generated function who's body
contains that statement.

For this work, we replaced a subset of the OpenMP implementation to generate
Tapir IR instead of vanilla LLVM IR with OpenMP runtime calls. Specifically, we
replace the two primary pragmas for task parallelism: \texttt{task\_spawn} and
\texttt{task\_wait}. By re-using code from Schardl et al. for code generation
of Cilk constructs, we were able to easily generate Tapir code for these OpenMP
pragmas. The ease with which this was completed is a testament to the quality of
the Tapir implementation.

While we did implement the codegen for the \texttt{task\_spawn} and
\texttt{task\_wait} constructs, it's worth noting here that even for these
pragmas, the implementation is incomplete. Currently, any clauses modifying the
behaviour of the pragmas is ignored. Probably the most common semantics this
will change of variable semantics, e.g. \texttt{shared} vs. \texttt{private}.
Surprisingly, this had little effect on the correctness of the entire Barcelona
OpenMP task suite. Indeed, it is likely that fixing this issue would increase
the performance of this work marginally, due to reducing the number of memory
copies for variables declared \texttt{private} in OpenMP code. This property
that behaviour wasn't changed significantly is worth revisiting, and we do so
in Section~\ref{Sec:Discussion}.

This work represented in this paper is only frontend implementation, which
means we use the only existing backend for Tapir. The existing backend
generates code that calls into the Cilk runtime, \texttt{libcilkrts}. This has
implications for adhering to the OpenMP specification. For example, environment
variables such as \texttt{OMP\_NUM\_THREADS} are ignored, replaced by
\texttt{CILK\_NWORKERS}.  We discuss some of these issues below, and return to
address others in the future work section.

There are other issues that had to be overcome to run full OpenMP programs
using Tapir as described above. Generally, in addition the the \texttt{task} 
and \texttt{taskwait} pragmas, a program has to contain initialization OpenMP
pragmas in order to run parallel code. A standard pattern for building
task-parallel OpenMP programs is to insert a \texttt{parallel} pragma to start
the necessary hardware threads, followed immediately by a \texttt{single}
pragma to have only one thread continue on the specified statement. Then the
other threads will get work from spawned tasks, instead of implicitly executing
the same code in a data parallel style. For the purpose of this paper, we took
the shortcut of replacing these pragmas with no-ops. This worked well for all
examples except for one, in which after the \texttt{parallel} and
\texttt{single} pragmas the top level task was called with an unnecessary
\texttt{task} pragma.  Because the wait was implicit and unhandled by our
implementation, the simple fix was to remove the \texttt{task} pragma. This was
the only change required to the source code. Our temporary no-op shortcut, of
course, doesn't follow the OpenMP specification, and should be addressed by
future work. See Section~\ref{Sec:Future} for a further discussion how how the 
implementation can be improved.

All code used for the implementation is available at \\
\texttt{https://github.com/lanl/openmpir-clang}. The version used for this
paper is tagged as LLVM17.

@@

To better understand how the OpenMP to Tapir compiler works, we turn to the
parallel fragment of the \texttt{fib} example from Section~\ref{Sec:OpenMP}. In
Figure~\ref{fig:tapir-example} we show what the input OpenMP code and the Tapir code
is generated.

@@

Upon entering the \texttt{else} branch, the code immediately detaches the basic
block corresponding to the call to \texttt{fib(n-1)}, labeled
\texttt{det.achd}. The continuation for the first call also immediately
detaches the basic block corresponding to the call of \texttt{fib(n-2)}.
Finally, the continuation for the second call immediately calls sync. This is
actually a nice example of code that will be optimized by Tapir. The LLVM code
shown in \figref{tapir-example} is not optimized. With optimizataions turned
on, Tapir will replace a \texttt{detach} of a function call followed
immediately by \texttt{sync} with a simple function call, resulting in faster
code.

@@

For evaluating the use of Tapir we compared performance on the Barcelona OpenMP Task 
Suite~\cite{barcelona} to several existing OpenMP implementations. The suite is a set 
of tests intended to test performance of OpenMP implementations
using both irregular and regular tasking.\footnote{An unbalanced tree search benchmark was
very recently added to the suite. Unfortunately we were unable to succesfully run 
it on any implementation, even for even medium input sizes. We have therefore left 
it out of our evaluation.  We will continue to work on this benchmark and will 
include the results if the paper is accepted to the workshop.} The following list provides abridged descriptions of 
each benchmark -- they are taken verbatim from~\cite{barcelona}:

@@

of a vector of n complex values using the Cooley-Tukey 

@@

\item \texttt{Fibonacci}: computes the n'th fibonacci number using a  recursive

@@

of a number of cells. The algorithm gets an input file with  cell’s
description  and  it  returns  the  minimum  area  size which includes all

@@

\item \texttt{Sort}: sorts a random permutation of n 32-bit numbers with  a

@@

so increase  the  task  granularity.  To  avoid  the overhead of  quicksort, an

@@

There are two classes of input sizes for the benchmarks. Some, like Floorplan,
require input files. For each of these, we chose the largest provided. For the
benchmarks with a parameter to adjust the size of the input, we attempted to 
choose a parameter size so that the fastest implementation was on the order of
seconds. The one exception was for \texttt{Fibonacci}. Because, as discussed
further below, the Intel implementation required large stacksize, we stopped at
42. The exact inputs used are:

@@

The Intel compiler required increasing the stacksize for \texttt{Fibonacci} and
\texttt{FFT}. The implementation seemed to use significantly more stack space
than the others, requiring setting \texttt{OMP\_STACKSIZE} to \texttt{256M} for
computing the 42nd Fibonacci number. \texttt{FFT} ran fine with an increase to
\texttt{16M}.

For other implementations, we compare to GCC 7.1, Clang 4.0.1, and Intel
17.0.0. The machine used is a two socket Intel Xeon E5-2683 v3 machine with
132GB of memory, running Linux 4.11.4. Each Xeon has 16 cores running at
2.1GHz, with 20M of L3 cache. All benchmarks were run using all 64
hyper-threads. 32 thread tests were run as a sanity check and showed
qualitatively similar results. It's worth noting that each of the other
implementations comes with it's own runtime. In this sense, performance is a
function of both the compiler and any optimizations it's able to perform, and
the runtime.  Understanding the interaction of these two parts is non-trivial,
as we will see when discussing results. Each compiler was run with the flags
\texttt{-O3 -fopenmp}. Our compiler was also run with \texttt{-ftapir} to
enable Tapir instructions to be lowered to a parallel runtime (in this case,
the default Cilk runtime).

>>>>

\title{OpenMPIR}
\subtitle{Implementing OpenMP tasks with Tapir}

\author{George Stelle}
\affiliation{\institution{Los Alamos National Laboratory}}
\email{stelleg@lanl.gov}

@@

\begin{abstract}
Optimizing compilers for task-level parallelism are still in their infancy.
This work explores a compiler front end that translates OpenMP tasking semantics to Tapir,
an extension to LLVM IR that represents fork-join parallelism. This enables
analyses and optimizations that were previously inaccessible to OpenMP codes,
as well as the ability to target additional runtimes at code generation. Using
a Cilk runtime back end, we compare results to existing OpenMP implementations.
Initial performance results for the Barcelona OpenMP tasking suite show
performance improvements over existing implementations.
\end{abstract}

@@

When writing task-parallel programs today, there is a large selection of
potential programming models and implementations to consider. Unfortunately, 
despite having some significant overlaps in semantics, parallel programming
models such as OpenMP~\cite{openmp}, Cilk~\cite{cilk}, Kokkos~\cite{kokkos},
HPX~\cite{hpx}, Charm++~\cite{charm}, Qthreads~\cite{qthreads},
pthreads\cite{pthreads}, MPI~\cite{mpi}, Chapel~\cite{chapel}, UPC~\cite{upc},
OpenCL~\cite{opencl}, and OpenACC~\cite{openacc}, among others, have limited ability to
interoperate either at the level of compile-time analysis or at run-time
execution. In addition, many of these programming models are implemented as
runtime libraries with only primitive compiler support. As a consequence, most are
not able to take advantage of potential compiler analysis and optimization
capabilities.  While there has been some recent work on specializing
compilers to reason about parallel programming libraries~\cite{Moss_2016},
fragmentation among models presents a significant challenge for compiler
writers, which must choose a single model, or multiple intermediate forms, for
analysis and optimization. An ideal solution would be a 
common representation of parallel semantics and constructs.

@@

Compilers implement the majority of their analyses and optimizations on
intermediate representations \textbf{(IRs)} which allow such transformations
to be written once with relative ease and apply to a variety of source
front-end languages such as C or C++.  Historically, the IR of mainstream
compilers such as LLVM~\cite{lattner2004llvm} or GCC's
Gimple~\cite{merrill2003generic} haven't suppored parallel constructs. As a
consequence, compilers using these IRs haven't had the ability to reason
about parallelism. The recent work of Schardl et al.\ has shown
that the LLVM IR can be extended to represent fork-join parallelism without
requiring a major rewrite~\cite{tapir}.  They did so by extending the LLVM instruction set
with three instructions capable of representing fork-join parallelism. To
benchmark performance and test accuracy they wrote a front end which translated
programs written with Cilk, a parallel programming model for C/C++, into the
Tapir IR. They also implemented a back end to \textbf{lower} Tapir programs to
vanilla LLVM IR with embedded Cilk runtime calls. They suggested that a similar
approach could be taken with OpenMP and other such frameworks.

In this work we put that suggestion to the test, implementing OpenMP tasks by
compiling them to Tapir instructions. This allows us to both analyze and optimize
OpenMP programs as well as compile programs written in OpenMP to use other runtimes.
For this paper our prototype implementation uses the Cilk runtime as a back end,
and we run our prototype implementation on the Barcelona OpenMP task suite~\cite{barcelona}. 
We discuss the kinds of optimizations this enables and how
the work can be extended to include other parallel constructs and semantics
in OpenMP. We also discuss the possibility of extending this work to other
programming models, which would help to alleviate some of the problems with fragmentation
previously discussed.

The contributions of this work include
\begin{itemize}
  \item A prototype front end that transforms OpenMP task constructs to Tapir IR;
  \item Benchmarks of the implementation using a Cilk runtime back end, showing
        performance improvements over existing OpenMP implementations; and
  \item An initial investigation into the reasons for these performance improvements.
\end{itemize}
Our hope is to move the community towards further conversations on a 
shared IR that can be used for many programming models. The work presented here only
represents a small step in that direction and there is still a herculean amount
of design and implementaton details to be considered. We argue that the benefits vastly outweigh 
the costs and we hope to see additional future work explore additional possibilities.

@@

The remainder of the paper is organized as follows.  In Section~\ref{Sec:Background} 
we give background and motivation for the presented work. Following that, we describe OpenMP
tasking in Section~\ref{Sec:OpenMP}, and Tapir in Section~\ref{Sec:Tapir}. We then 
discuss the implementation, and how we compile OpenMP task constructs to Tapir IR in
Section~\ref{Sec:Implementation}. In Section~\ref{Sec:Evaluation} we describe the 
evaluation setup, including which implementations we compare to and how. We discuss
the results of evaluation in Section~\ref{Sec:Results}, including possible
explanations for discrepancies. We discuss the implications of the results and the
significance of this work in a large context in~\ref{Sec:Discussion}. Finally,
we cover some of the many ways the work could be extended in
Section~\ref{Sec:Future}, and conclude in Section~\ref{Sec:Conclusion}.

@@

LLVM has gained traction in the broad community for its simplicity,
relative ease of both experimentation and modifications, and extensibility~\cite{lattner2004llvm}.
Historically, LLVM has only had a sequential semantics: there is no explicit notion
of concurrent or parallel processes. This has resulted in existing compilers
treating parallel constructs as thin wrappers for calls into run-time
systems. These calls are infeasible for the compiler to reason about because of
the complexity and flexibility of the runtimes being called, and the challenge
of reasoning about the corresponding semantics in a uniform fashion. The
addition of parallel constructs were recently explored by Schardl et
al.~\cite{tapir}. By adding three first-class instructions to LLVM, Tapir
enables analyses and optimizations of parallel constructs. We briefly describe
the key aspects of Tapir in Section~\ref{Sec:Tapir}.
% I removed this as the reference was already given...   PM For a more complete
% description see \cite{tapir}.

Arguably the most common parallel programming model for on-node parallelism
in HPC, OpenMP has grown from being a method for implementing parallel loops into a
large and complex specification for parallelism. This paper focuses on a 
relatively recent addition to the OpenMP specification: the \texttt{task}
and \texttt{taskwait} constructs. A detailed description of these constructs
is presented in Section~\ref{Sec:OpenMP} and a more complete description can 
be found in the OpenMP specification~\cite{openmp}.

@@

In this section we further discuss a significant challenge in HPC:
fragmentation of parallel programming models. There are many ways to write
parallel programs~\cite{openmp, qthreads, chapel, cilk, kokkos, legion, upc,
mpi}.  From libraries, to language extensions, to standalone languages, to
combinations of the these, the fragmentation of parallel programming models is a
problem that is only getting worse. This means that when writing tooling,
optimizations, or analyses, one much choose which model to target as well as
understand points of contention and interoperability between them. A common
difficulty among all of these is that reasoning about parallel programs is hard
and compilers can potentially help reduce this burden. This is exactly
analogous the problem LLVM solves: compiler optimizations are hard, so sharing
them is valuable. Tapir is a demonstration that the the LLVM approach to
compilation of serial code can be effectively extended to help reason about
parallel code.

With fragmentation comes questions of compatibility or composability of 
different programming models. To illustrate this, consider the example of
a program using a library that uses OpenMP to implement parallelism internally. 
If this program uses a different programming model for parallelism, e.g.\ Cilk,
there is currently no good solution for ensuring that the two different back ends
will cooperate with management of hardware resources. 
%
Addressing these concerns is a major motivation for our use of Tapir.  By
standardizing an IR, in addition to easing the development of optimizations,
one avoids duplicating work across different implementations. It also enables
the possibility of multiple programming models coexisting peacefully. We will
return to this consideration later in Section~\ref{Sec:Discussion} when discussing
future work.

\section{Tapir} \label{Sec:Tapir}
Tapir is an extension to LLVM proposed by Schardl, Moses, and Leiserson
seeking to resolve issues in optimizing parallel code. Prior to
the introduction of Tapir compilers would represent parallel programs
by directly translating their syntax to opaque runtime calls. This allowed
compilers to support parallel programs but meant that traditional 
optimizations such as code motion~\cite{} or common-subexpression elimination~\cite{}
weren't able to reason about parallel programs. This often led to parallel
programs running significantly slower than expected. In the Tapir project
the authors show that it is possible to represent and optimize fork-join
parallel programs with relatively minimal modifications to the compiler, namely
the introduction of three instructions designed to interface well with an
existing compiler.

@@

Consider the first code segment defined in~\figref{search},
a simple OpenMP program which performs divide-and-conquer search.
However, as written, there is a simple optimization that can be performed,
that the value \code{(low+high)/2} may be computed once before either
parallel task, thereby avoiding redundant computations. For
the serial version of this program optimizations like this are performed
automatically. However, without the use of Tapir, these sorts of optimizations
are not run on parallel programs.

@@

\subsection{Tapir Internals}
Tapir is a compelling solution to this problem because it allows existing
analysis and optimizations to work on parallel programs without significant modifications.
To properly describe Tapir requires a brief introduction to key aspects of LLVM.
LLVM represents expression/evaluation order through a control-flow graph (CFG) of blocks that contain 
sequences of instructions. Blocks always end with terminator instructions that
define the edges of the control flow graph. For example, \figref{CFG} shows a simple
function and its corresponding LLVM code.

@@

To represent parallelism Tapir introduces the \code{detach}, \code{reattach}, and
\code{sync} instructions. The \code{detach} statement is syntactically similar to
a conditional branch in that it is a terminator instruction with with two blocks.
However, unlike the conditional branch which denotes a choice between successors, the \code{detach} instruction denotes that its two successors may run in parallel. Semantically,
the first successor of a \code{detach} instruction is referred to as the \textbf{detached block} and represents a task that can, but is not required to, run in parallel with the
\textbf{continuation block}, or the second successor to the detach statement. The \code{detach}
is analagous to a \code{fork} in most fork-join models.

The end of a task created by a \code{detach} instruction must end with a \code{reattach} instruction to the corresponding continuation block. The \code{reattach} instruction
is used to denote the end of a parallel task.
%
Finally, one may use the \code{sync} instruction to wait for any outstanding tasks created
by \code{detach} instructions within the current function to complete. The \code{sync}
instruction is analagous to a \code{join} in most fork-join frameworks.

@@

The initial design of the OpenMP task model established 
a basic framework for asynchronous task parallel execution in OpenMP programs~\cite{ayguade09design}.
Subsequent versions of the OpenMP specification, up to the current version 4.5, 
have added new features to the tasking model.  The \texttt{depend} 
clause codifies data dependences among tasks, indicating that a data location 
is an input or output of a task.  The runtime system ensures that a task is 
not scheduled until its input dependencies are fulfilled.  The \texttt{taskloop} 
construct combines groups of independent loop iterations into explicit tasks, 
enabling composition of concurrent loop execution and independent explicit 
tasks within the same OpenMP parallel region. The \texttt{taskgroup} construct 
waits on not only all child tasks, but all descendent tasks, providing a deep 
synchronization. The \texttt{taskwait} construct allows the application to 
indicate a point at which the implementation may suspend the current task to 
work on other tasks, as may be desired for long-running tasks that generate 
many others.  The task concept was also leveraged to provide asynchronous 
offload of data and computation to accelerators by applying the \texttt{nowait} 
clause to device constructs such as the \texttt{target} construct to generate an 
asynchronous \textit{target task}.

@@

Tapir is implemented as an extension to the LLVM instruction set. Clang is a C
family compiler that has support for OpenMP extensions and targets
LLVM~ \cite{clang}. The existing Clang OpenMP implementation maps OpenMP constructs directly
to OpenMP runtime library calls by wrapping a C statement in a
\texttt{CapturedStmt}. This replaces a C statement with a function call into
the OpenMP runtime library, along with a machine generated function whose body
contains that statement.

For this work we replaced a subset of the OpenMP implementation to generate
Tapir IR instead of vanilla LLVM IR with OpenMP runtime calls. Specifically, we
replace the two primary \texttt{task\_spawn} and
\texttt{task\_wait} pragmas for task parallelism. By re-using code from Schardl et al.\ for code generation
of Cilk constructs, we were able to easily generate Tapir code for these OpenMP
pragmas. The ease with which this was completed is a testament to the quality of
the Tapir implementation.

While we did implement the codegen for the \texttt{task\_spawn} and
\texttt{task\_wait} constructs, it's worth noting that even for these
pragmas the implementation is incomplete. Currently, any clauses modifying the
behavior of the pragmas is ignored. Probably the most common semantics this
will change is for the semantcis of variables, e.g.\ \texttt{shared} vs.\ \texttt{private}.
Surprisingly, this had little effect on the correctness of the entire Barcelona
OpenMP task suite. Indeed, it is likely that fixing this problem would marginally increase
performance by reducing the number of memory
copies for variables declared \texttt{private} in OpenMP code.  The fact
that behaviour wasn't changed significantly is worth revisiting, and we do so
in Section~\ref{Sec:Discussion}.

The work represented in this paper is only front-end implementation, which
means we use the only existing back end for Tapir. The existing back end
generates code that calls the Cilk runtime \texttt{libcilkrts}. This has
implications for adhering to the OpenMP specification. For example, environment
variables such as \texttt{OMP\_NUM\_THREADS} are ignored and replaced by
\texttt{CILK\_NWORKERS}.  We discuss some of these issues below, and return to
address others in the discussion of future work.

There are other issues that had to be overcome to run full OpenMP programs
using Tapir. Generally, in addition the the \texttt{task} 
and \texttt{taskwait} pragmas, a program must contain pragmas to initialize OpenMP
%pragmas in order to run parallel code.
A standard pattern for building
task-parallel OpenMP programs is to insert a \texttt{parallel} pragma to start
the necessary hardware threads, followed immediately by a \texttt{single}
pragma to have only one thread continue on the specified statement. Then the
other threads will get work from spawned tasks instead of implicitly executing
the same code in a data parallel manner. For the purpose of this work we took
the shortcut of replacing these pragmas with no-ops. This worked well for all
examples except for one in which after the \texttt{parallel} and
\texttt{single} pragmas the top level task was called with an unnecessary
\texttt{task} pragma.  Because the wait was implicit and unhandled by our
implementation, the simple fix was to remove the \texttt{task} pragma. This was
the only change required to the source code. Our temporary no-op shortcut, of
course, doesn't follow the OpenMP specification, and should be addressed by
future work. Section~\ref{Sec:Future} discusses how how the 
implementation can be improved.

All code used for the implementation is available at \\
\texttt{https://github.com/lanl/openmpir-clang}. The version used for this
paper is tagged LLVM17.

@@

To better understand how the OpenMP to Tapir compiler works, we turn to the
parallel fragment of the \texttt{fib} example from Section~\ref{Sec:OpenMP}. In
Figure~\ref{fig:tapir-example} we show the input OpenMP code and the Tapir code
that is generated.

@@

Upon entering the \texttt{else} branch, the code immediately detaches the basic
block corresponding to the call to \texttt{fib(n-1)}, labeled
\texttt{det.achd}. The continuation for the first call also immediately
detaches the basic block corresponding to the call of \texttt{fib(n-2)}.
Finally, the continuation for the second call immediately calls sync. This is
a nice example of code that will be optimized by Tapir. The LLVM code
shown in \figref{tapir-example} is not optimized. With optimizations turned
on Tapir will replace a \texttt{detach} of a function call followed
immediately by \texttt{sync} with a simple function call, resulting in faster
code.

@@

To evaluate the use of Tapir we compared performance on the Barcelona OpenMP Task 
Suite to several existing OpenMP implementations. The suite is a set 
of tests intended to test performance of OpenMP implementations
using both irregular and regular tasking.\footnote{An unbalanced tree search benchmark was
very recently added to the suite. Unfortunately we were unable to succesfully run 
it on any implementation, even for medium input sizes. We have therefore left 
it out of our evaluation.  We will continue to work on this benchmark and will 
include the results if the paper is accepted to the workshop.} The following list provides abridged descriptions of 
each benchmark taken verbatim from the test suite distribution.

@@

of a vector of $n$ complex values using the Cooley-Tukey 

@@

\item \texttt{Fibonacci}: computes the $n$th Fibonacci number using a  recursive

@@

of a number of cells. The algorithm gets an input file with  cells'
descriptions  and  it  returns  the  minimum  area  size which includes all

@@

\item \texttt{Sort}: sorts a random permutation of $n$ 32-bit numbers with  a

@@

to increase  the  task  granularity.  To  avoid  the overhead of  quicksort, an

@@

There are two classes of input sizes for the benchmarks. Some, like Floorplan,
require input files. For each of these we chose the largest provided. For the
benchmarks with a parameter to adjust the size of the input, we attempted to 
choose a parameter size so that the fastest implementation ran on the order of
seconds. The one exception was for \texttt{Fibonacci}. Because, as discussed
further below, the Intel implementation required a large stack size, we stopped at
42. The exact inputs used are:

@@

The Intel compiler required increasing the stack size for \texttt{Fibonacci} and
\texttt{FFT}. The implementation seemed to use significantly more stack space
than the others, requiring setting \texttt{OMP\_STACKSIZE} to \texttt{256M} for
computing the 42nd Fibonacci number. \texttt{FFT} ran with an increase to
\texttt{16M}.

For other implementations, we compare to GCC 7.1, Clang 4.0.1, and Intel
17.0.0. The machine used is a two socket Intel Xeon E5-2683 v3 machine with
132GB of memory running Linux 4.11.4. Each Xeon has 16 cores running at
2.1GHz, with 20M of L3 cache. All benchmarks were run using all 64
hyper-threads. 32-thread tests were run as a sanity check and showed
qualitatively similar results. It's worth noting that each of the other
implementations comes with its own runtime. In this sense, performance is a
function of both the compiler and any optimizations it is able to perform, and
the runtime.  Understanding the interaction of these two parts is non-trivial
as we will see when discussing results. Each compiler was run with the flags
\texttt{-O3 -fopenmp}. Our compiler was also run with \texttt{-ftapir} to
enable Tapir instructions to be lowered to a parallel runtime (in this case,
the default Cilk runtime).

As mentioned in the Section~\ref{Sec:Implementation}, the gaps in the implementation
changed program behavior in a couple cases. In the case of \texttt{FFT}, we were
forced to remove an unecessary \texttt{task} pragma, as the current implementation
doesn't insert the implicit barrier at the end of a \texttt{parallel} region. This 
was a one line fix in the benchmark, and would be fixed properly by handling
the \texttt{parallel} and \texttt{single} pragmas correctly in our implementation.
